---
title: "Arrest of the First Seven Days of October 2024 and Correctional Facilities on O'ahu"
author: "Pamela Oda"
date: "10-20-2024"
output: 
  html_document:
    theme: flatly
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("vembedr")
```

As someone that is pursuing a Bachelors of Science in Criminology and Criminal Justice and Data Science, Analytics and Visualization, I hope to pursue a career as a research analyst within the State of Hawai'i criminal justice system, crime analyst or intelligence analyst. In addition, being born and raised in Hawai'i, I wanted to dedicate my project towards something that I think would be valuable to my community. 

## Loading Libraries 
To start, let's load the necessary libraries to create the maps:
```{r}
library(rinat)
library(sf)
library(tidyverse)
library(leaflet)
library(osmdata)
library(plotly)
library(readxl)
```

## Creating the Bounding Box of O'ahu
Now that the necessary libraries are loaded, let's start by creating the bounding box of O'ahu. To create the bounding box, I will use the `osmdata` package function `getbb()` to get the bounding box.
```{r}
oahu_bb <- getbb("Oahu")
oahu_bb
```

## Retriving Geographical Data
Now, with the bounding box and the `osmdata` package, we will retrieve the streets, small streets, coastline, national parks and the protected areas of the Island of O'ahu. This will give us a better idea of the geographical structure of the island.
```{r}
#retrieving data of streets on Oahu
oahu_streets <- oahu_bb %>%
  opq() %>%
  add_osm_feature("highway", c("motorway", "primary", "secondary", "tertiary")) %>%
  osmdata_sf()

#retrieving data of small streets on Oahu
oahu_small_streets <- oahu_bb %>%
  opq() %>%
  add_osm_feature(key = "highway", value = c("residential", "living_street", "unclassified", "service", "footway")) %>%
  osmdata_sf()

#retrieving data of coastline of Oahu
oahu_coast <- oahu_bb %>%
  opq() %>%
  add_osm_feature(key = "natural", value = "coastline") %>%
  osmdata_sf()

#retrieving data of national parks in Oahu
oahu_np <- oahu_bb %>%
  opq() %>%
  add_osm_feature(key = "boundary", value = "national_park") %>%
  osmdata_sf()

#retrieving data of protected areas in Oahu
oahu_protected <- oahu_bb %>%
  opq() %>%
  add_osm_feature(key = "boundary", value = "protected_area") %>%
  osmdata_sf()
```

## Creating a Static Map of the Island of O'ahu
Now that we have the geographical data, let's create the first map of the Island of O'ahu.
```{r}
p <- ggplot() +
  geom_sf(data = oahu_streets$osm_lines, inherit.aes = FALSE, color = "#ffbe7f", size = .4, alpha = .8) +
  geom_sf(data = oahu_coast$osm_lines, inherit.aes = FALSE, color = "black", size = .8, alpha = .5) +
  geom_sf(data = oahu_np$osm_polygons, inherit.aes = FALSE, color = "brown", size = .2, alpha = .8) +
  geom_sf(data = oahu_protected$osm_polygons, inherit.aes = FALSE, color = "green", size = .2, alpha = .8) +
  geom_sf_text(size = 1, data = oahu_protected$osm_polygons, aes(label = name)) + 
  coord_sf(xlim = c(-158.28058, -157.64863), ylim = c(21.25482, 21.71201), expand = TRUE) + 
  ggtitle("Island of O'ahu", subtitle = "Based on iNaturalist Data as of October 2024") +
  theme_bw() +
  labs(
    color="Common Name",
    x = "Longitude",
    y = "Latitude"
  ) 

p
```
As you can see in the figure above, I created a map that outlines the various geographical features of the Island of O'ahu. The data that was used to create this map came from iNaturalist of the current data that it has of the island. This static map provides the foundation of what the island looks like, highlighting the streets, protected areas, national parks, and so on to give a full picture of the island. 

Unfortunately, being that there are many streets on the island, the map does become clustered when you look at certain areas of the map, making it hard to differentiate the different streets. 
On the other hand, this map does provide the locations of national parks and protected areas where there are native plants and species. With that said, law enforcement such as the Honolulu Police Department can work with the different organizations in creating protection plans for these areas to ensure the protective of the native plants and species. 

## Creating the October 2024 Arrest Variable
Now, let's take the necessary steps towards making an interaction map of the Island of O'ahu by creating the variable "October2024_Arrests" and assigning the data set that contains the longitude and latitude of the arrests for the first seven (7) days of October 2024. 

Let's read in the Excel file using the `read_excel()` function and assign the dataset to the variable `October2024_Arrests`. 
```{r}
October2024_Arrests <- read_excel("~/Downloads/O'ahu Arrests in October 2024.xlsx")
View(October2024_Arrests)
```

## Craeting an Interactive Map of the Arrests Locations on the Island of O'ahu
Now, let's create an interactive map of the arrests locations within the first seven (7) days of October 2024 using the leaflet function.
```{r}
leaflet() %>%
  addTiles() %>%
  addCircles(
    lng = October2024_Arrests$Longitutde,
    lat = October2024_Arrests$Latitude,
    label = October2024_Arrests$Location,
    color = "red"
  )
```
Normally with leaflet, you use the polygons feature, but here, we are just trying to display the specific location of the arrest using. Each red dot represents an arrest that was made during the first seven (7) days in October of 2024 according to the Honolulu Police Department (HPD) Arrest Log. 

Through this map, you are able to visually see where arrests are most common on the island as well as the frequency of arrests based on the seven day period. This map also displays the areas where a crime or crimes have taken place recently around the island. More specifically, according to the map, majority of the crimes that occurred within this seven (7) day period were in the Honolulu area including Chinatown, Ala Moana, Waikiki and Downtown. 

Unfortunately, not all of the arrests that occurred within this seven (7) day period is recorded on this map. As I was manually creating the spreadsheet of this data set and utilizing a website to convert the location provided on the arrest logs to longitude and latitude points, the website could not recognize some of the locations. With that said, those arrest were excluded from the data set. 

Ideally, maps like these, whether they are produced weekly or daily would be useful not only to law enforcement when strategizing patrol but also the community as they would be able to uses these maps as an educational tool in being aware of their surroundings and being vigilant. These maps could potentially contribute to lower crime rates in visualizing the areas of concern for law enforcement as well as the safety of individuals in the community due to more awareness of what is happening in their community. 

## Creating the Correctional Facilities Variable
Now, let's take the necessary steps towards making this next interaction map of the Island of O'ahu by creating the variable "Oahu_Corrections" and assigning the dataset that contains the longitude and latitude of the correctional facilities.

Let's read in the Excel file using the `read_excel()` function and assign the dataset to the variable `Oahu_Corrections`. 
```{r}
Oahu_Corrections <- read_excel("~/Downloads/O'ahu Correctional Facilities.xlsx")
View(Oahu_Corrections)
```

## Craeting an Interactive Map of the Correctional Facilities on the Island of O'ahu
Now, let's create an interactive map of the correctional facilitates on the Island of O'ahu.
```{r}
leaflet() %>%
  addTiles() %>%
  addCircles(
    lng = Oahu_Corrections$Longitude,
    lat = Oahu_Corrections$Latitude,
    label = Oahu_Corrections$Facility,
    color = "blue"
  )
```
In the interactive map above, you can see the six correctional facilities across O'ahu. This data set was created manually through collecting the information of the correctional facilities and converting the location to longitude and latitude points for plotting purposes. Each blue dot represents one of the correctional facilities and if you hover over the blue dot, it will indicate which facility it is. 

Unfortunately, as there are only six correctional facilities on the Island of O'ahu, the blue dots do blend into the map, making it hard to identify when looking at the map above.

Although there are only six facilities, the Honolulu Police Department (HPD) can use this map to work with the Department of Corrections in developing a partnership around these areas for further safety and security around these facilities and areas around the island. 

## Craeting an Interactive Map of the Arrest of the First Seven Days of October 2024 and the Correctional Facilities on the Island of O'ahu
Now, let's create an interactive map of the correctional facilties on the Island of O'ahu.
```{r}
leaflet() %>%
  addTiles() %>%
  addCircles(
    lng = October2024_Arrests$Longitutde,
    lat = October2024_Arrests$Latitude,
    label = October2024_Arrests$Location,
    color = "red"
  ) %>%
  addCircles(
    lng = Oahu_Corrections$Longitude,
    lat = Oahu_Corrections$Latitude,
    label = Oahu_Corrections$Facility,
    color = "blue"
  )
```
As displayed in the figure above, I combined the arrest data and the correctional facilities data to create an overlapping interactive map. 

With this map, you are able to see the locations of the six correctional facilities on the Island of O'ahu along with the arrest incidents that occurred during this seven (7) day period. As you can see, not many of the crimes are near these facilities besides the Oahu Community Correctional Center, which is located in the Honolulu areas where the majority of the crimes occurred. You'll also notice that there are a handful of crimes that occurred on the west side of the island, but there is no correctional facility on that side of the island.

With that said, these maps could contribute to the development of more correctional facilities on the island as overcrowding in the current facilities is an issue that the Hawai'i Department of Corrections currently faces. Also, as stated in the previous maps, these maps would be a useful tool in bringing awareness to the community of the crimes that occur around the island as well as valuable information for law enforcement when on patrol. More importantly, these maps will also display the areas of concern on the island and what areas need the most resources and help in keeping the community safe.

All in all, as someone that wants to contribute towards making our community safer, using data science to conduct research and studies can help our community in making data driven decisions in making our community safer. 